Limited physiological plasticity and reduced relative variance in physiological rates of ectotherm populations under climate change

### 2024-07-06 ms.qmd
Fixing up discussion of thermal variability and also life stages

Reversible phenotypic plasticity, such as physiological acclimation, is relatively rapid and can be fine-tuned to proximate environmental conditions making it the first ‘line-of-defense’ against rapid environmental change [@Dewitt1998; @Scheiner2020]. For example, physiological rates are known to speed up as temperature increases because of the thermodynamic effects on chemical reaction rates -- so called 'acute' temperature responses (@fig-concept). Acute temperature increases that push organisms from their thermal optimum can be mitigated by plastically adjusting reaction rates in response to chronic temperatures through acclimation [@Seebacher2015-wi; @Havird2020-mj]. Acclimation alters acute thermal sensitivity to offset the potentially negative effects of acute temperature changes (e.g., higher energetic demands). Acclimation, however, does not normally result in physiological rates returning to their original state (‘complete’ compensation - *sensu* @huey1999testing), but rather are only partially compensated [@huey1999testing; @Havird2020-mj].

Phenotypic plasticity is expected to evolve in populations when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Thermal variation and predictability are known to differ across habitats (terrestrial, marine and freshwater) [@steele2019comparison], and as such, we expect the extent of plasticity of organisms within these habitats to vary. Species occupying terrestrial habitats often show weak acclimation abilities, and so, are likely to be particularly vulnerable to climate change given their greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper; @morley2019physiological]. Marine and freshwater habitats appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022].

Physiological acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to allow organisms to maintain physiological performance around a fitness optimum despite changes in the environment [@seebacher2019histone; @little2013thyroid; @taff2016endocrine]. The focus of research up to now has been primarily on mean physiological responses neglecting how variability in physiological rates might also be impacted at higher temperatures. As mean physiological rates increase with temperature there is reason to believe that the intrapopulation variability might also be impacted. NEED MORE ABOUT WHY VARIANCE NOT JUST IMPORTANCE

@@

#### *Calculating acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates*
As indicated above, $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ can all be taken from samples of organisms measured acutely at two temperatures or after having been acclimated these same temperatures. For studies that measure acute and acclimated responses we used the mean, standard deviation, and sample size to derive both acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates. For studies that only measured $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ acutely we could only calculate acute versions of these effect size estimates. For all effect sizes the higher temperature was in the numerator and the lower of the two temperatures in the denominator. As such, positive effect sizes suggest that the mean or relative variance is larger at the higher of the two temperatures (numerator) when standardized to 10&deg;C.

#### *Assumptions of $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates*

$lnRR_{Q_{10}}$, $lnCVR_{Q_{10}}$ and $lnVR_{Q_{10}}$, as with $Q_{10}$ more generally, all assume that the effect of temperature on physiological rates (or changes in relative variance) is log-linear. While this is likely in our data given that we restricted our analysis of $Q_{10}$ to standard operating temperatures for a given species, it may not always be satisfied given the diversity of species in our dataset. $Q_{10}$ [@van1884etudes] has been used exstensively in the physiological literature to successfully address a multitude of questions [e.g., @Havird2020-mj; @Seebacher2015-wi]. However, there is a preference for using a Boltzmann – Arrhenius (BA) relationship (or its more recent extension, the Sharpe-Schoolfield model [@molnar2017thermal; @michaletz2024hotter]) to model thermal effects on physiological rates [@gillooly2001effects; @michaletz2024hotter]. While debate still exists over the utility of $Q_{10}$ when modelling temperature-dependence it is important to recognise that both BA and $Q_{10}$ can exhibit curvilinearity as temperatures increase (as discussed in [@michaletz2024hotter]). @white2012information also showed that the BA model may not always perform better. For example, in eukaryotes, modelling thermal dependence using $Q_{10}$ provided a 5.8-fold better fit to metabolic rate data than the BA relationship [@white2012information]. Given that studies included in our analysis never measured full performance curves at acute and acclimation temperatures it was not possible for us to compare different models of thermal dependence. Nonetheless, $Q_{10}$-based effect sizes remain the most practical effect-size for comparing thermal dependence when using existing empircial data, with the benefit that these effects having convenient properties that make them suitable for meta-analysis. Nonetheless, we control for possible violations of the log-linearity assumption in our analyses (see below).

@@

In addition to acclimation, all our models corrected for the possibility of the log-linearity assumption of our effect sizes being violated. We predicted that, if $lnRR_{Q_{10}}$, $lnCVR_{Q_{10}}$ and $lnVR_{Q_{10}}$ were not strictly log-linear than we would see a decrease in average effect size for studies applying higher temperature treatments because these temperatures are expected to either converge on or cross the thermal optima of the performance curve causing reaction rates to decelerate or decrease beyond $T_{op}$ [@michaletz2024hotter]. The benefit of such an approach means that we could still take advantage of the excellent properties of $Q_{10}$ as an effect size while statistically correcting for the potential non-linearity that would be expected in the data at high treatment temperatures. Given that our data included a wide range of species and habitats, we also included a random slope of maximum temperature that varied across species because we, *a priori*, expected that each species will vary in their thermal performance curve and the chosen temperatures experimentally manipulated along the performance curve. To improve model interpretation, we also mean-centered the maximum temperature and included it in our models.

@@

Changes in relative variance in physiological rates across habitats was nearly equivalent (marine) or larger (terrestrial) compared to their capacity to acclimate. Only in freshwater ectotherms did acclimation in physiological rates coincide with no change in relative variation. Mechanistically, it is unclear what is contributing to the reduction in relative variation in physiological rates at higher temperatures. One possible explanation may be that biochemical reactions are operating more efficiently or are relying on more conformationally stable enzymes at higher temperatures impacting between individual variation [@tattersall2012coping; @angilletta2009thermal; @somero1995proteins]. Such a mechanism may also help explain the differences in relative variance across habitat types. For example, terrestrial ectotherms may rely on more specialised enzymatic variants that operate over a more limited temperature range individually but together allow for organisms to function over a wider thermal range necessary to operate in more variable thermal environments [@angilletta2009thermal]. Acclimation responses in marine and freshwater organisms would also result in the maintenance, or even increase, in variance at higher temperatures due to the expression of alternative enzymes that function more effectively at higher temperatures. Given that biochemical traits in marine and freshwater organisms acclimated, and their relative variance increased in freshwater after acclimation, this seems plausible.

@@

### ***Plasticity and relative variance in physiological rates do not differ between life stages***
Differences in plasticity is expected between different life-stages because of distinct patterns of mortality, habitat use and behaviour between early and late life history stages [@martin2015age; @stearns1976life]. Age has also been shown to reduce the capacity for acclimation [e.g., see @rossi2019ageing]. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults while @rossi2019ageing show that older air-acclimated fish had a diminished scope for gill remodeling relative to younger fish in mangrove rivulus (*Kryptolebias marmoratus*).

Despite these examples, more broadly, we show little difference between early and late life acclimation capacities and changes in relative variance across habitats in our data indicating that such responses are likely context or trait dependent [@carter2020life; @moghadam2019heat]. The lack of differences we observed may also be due to the fact that both juvenile and adult animals occupy similar ecological niches with similar dispersal and thermoregulatory behaviors making physiological responses to temperature similar. It is possible that temperatures experienced earlier in development, for example egg or larval stages, exhibit greater capacity for acclimation because of limited dispersal capabilities and very different habitats. Early developmental constraints are known to affect how physiological systems respond later in life because of programmatic effects [@angilletta2009thermal; @noble2018developmental; @o2019developmental], and as such, developmental conditions may also cascade to . Such effects would manifest in juveniles life-history stages

### ***No signature of past climate on capacity for physiological plasticity***
Theoretical models predict that plasticity should evolve in populations experiencing greater environmental variability (spatial or temporal), particularly when oscillations are predictable over time to make environmental cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogeneity in terrestrial habitats [@steele2019comparison] might then suggest that plasticity should be more likely to evolve in terrestrial environments. However, if thermal variability is too high and unpredictable, the rates of acclimation slow, and there are costs associated with re-modelling physiological processes [which seem likely, @angilletta2009thermal] then we would instead expect canalised responses [@angilletta2009thermal; @Seebacher2015-wi; @loughland2020differences]. The lack of acclimation in terrestrial ectotherms we observed is consistent with the latter hypothesis, and is supported by other meta-analyses of heat tolerance [@barley2021limited; @Gunderson2015-hz] (which we did not consider here). However, our results do contrast with comparative studies that show partial support for environmental variability being related to acclimation responses in aquatic habitats [@Seebacher2015-wi; ]

Modelling correct microenvironments for over 200 species across such diverse habitats is challenging. The lack of a relationship between physiological plasticity and environmental variability and predictability could be due to the fact that ectotherms simply seek out microhabitats that make their environment quite stable [@huey2012predicting].  Therefore, it is possible that more fine-scale resolution of the habitats that organisms occupy will elucidate expected patterns. Randomly fluctuating environments have been suggested to select for reduced plasticity, particularly if plastic responses are costly and this has been demonstrated by a number of empirical studies [@Leung2020; @rescan2022experimental; @leung2023phenotypic]. For example, using seed beetles (*Callosobruchus maculatus*), @hallsson2012selection showed that experimentally evolving populations under randomly fluctuating thermal conditions did not show any increase in plasticity, but had reduced plasticity. @Leung2020 also showed reduced plasticity in morphological traits of experimentally evolving algae populations (*Dunaliella salina*) when environments were less predictable [see also @leung2023phenotypic]. While it is unclear how these findings relate directly to acclimation because they integrate multiple forms of plasticity, they do suggest that there are costs to being plastic or that the environmental signals are insufficient to trigger endocrine and epigenetic mechanisms that lead to plasticity when environments are not predictable [@Leung2020; @hendry2016key]. The fact that aquatic ectotherms did show a capacity to acclimate suggests that environmental predictability, which is expected to be higher in aquatic environments, may be more important than environmental variability in driving the evolution of plasticity, but it is unclear how this can be measured across diverse taxa at a biologically relevant spatial and temporal scale.

@@

Enhanced knowledge of how variation in physiological rates vary across populations and species and the degree to which they can be adjusted in response to the environment may lead to more informed predictions about the ecological and evolutionary dynamics of natural populations [@forsman2015rethinking; @seebacher2023can; @cooke2021one; @Sanderson2023]. While we show general patterns across taxa and habitats it is important to recognise that this does not mean such patterns apply to all populations. Substantial variation in acclimation responses and changes in variance exist among populations, as evidenced by wide prediction intervals. Nonetheless, changes in the relative variance in physiological rates could be better incorporated into physiological and ecological models to provide more nuanced, and possibly more realistic, predictions about the impacts of climate change on natural populations. While we do not yet understand the relative contribution of environmental and genetic factors to variance changes, models could better decouple how different levels of heritability and total variance impact evolutionary and ecological predictions. Our meta-analysis now provides the opportunity to parameterise such models, and ensure they are better aligned with empirical findings.

Many fascinating questions remain unanswered that will require greater focus on the consequences of changes in variance (rather than just the mean). Particularly interesting questions include: How does a reduction in physiological rate variance change energy flow across trophic levels within communities? What are the biochemical, cellular, and physiological mechanisms that underlie reduced variance in physiological rates? Are reductions in variance in one trait associated with reductions in correlated traits, or do some traits increase while others decrease? How much of the reduction in variance is driven by lower levels of genetic variance? Answers to these questions will require integrative approaches that combine empirical and theoretical work across multiple levels of biological organisation but will likely provide useful advances in understanding the full consequences that climate change will have on ectotherms across all major ecosystems.

@@

>>>>

Reversible phenotypic plasticity, such as physiological acclimation, is relatively rapid and can be fine-tuned to proximate environmental conditions making it the first ‘line-of-defense’ against environmental change [@Dewitt1998; @Scheiner2020]. For example, physiological rates are known to speed up as temperature increases because of the thermodynamic effects on chemical reaction rates -- so called 'acute' temperature responses (@fig-concept). Acute temperature increases that push organisms from their thermal optimum can be mitigated by plastically adjusting reaction rates in response to chronic temperatures through acclimation [@Seebacher2015-wi; @Havird2020-mj]. Physiological acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to allow organisms to maintain physiological performance around a fitness optimum despite changes in the environment [@seebacher2019histone; @little2013thyroid; @taff2016endocrine]. Acclimation alters acute thermal sensitivity to offset the potentially negative effects of acute temperature changes (e.g., higher energetic demands). Acclimation, however, does not normally result in physiological rates returning to their original state (‘complete’ compensation - *sensu* @huey1999testing), but rather are only partially compensated [@huey1999testing; @Havird2020-mj].

Reversible forms of phenotypic plasticity, such as acclimation, are expected to evolve in populations when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Thermal variation and predictability are known to differ across habitats (terrestrial, marine and freshwater) [@steele2019comparison], and as such, we expect organisms within these habitats to vary in their capacity for plastic adjustments in physiology. Species occupying terrestrial habitats have been shown to exhibit weak acclimation abilities, and so, are likely to be particularly vulnerable to climate change given their greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper; @morley2019physiological]. In contrast, marine and freshwater habitats appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, the focus of research up to now has been primarily on mean physiological responses neglecting how variability in physiological rates might also be impacted at higher temperatures.

As mean physiological rates increase with temperature there is reason to believe that the intrapopulation variability might also be impacted. NEED MORE ABOUT WHY VARIANCE NOT JUST IMPORTANCE

@@

#### *Calculating acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates*
As indicated above, $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ can all be taken from samples of organisms measured acutely at two temperatures or after having been acclimated these same temperatures. For studies that measure acute and acclimated responses we used the mean, standard deviation, and sample size to derive both acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates. For studies that only measured $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ acutely we could only calculate acute versions of these effect size estimates. For all effect sizes the higher temperature was in the numerator and the lower of the two temperatures in the denominator. As such, positive effect sizes suggest that the mean or relative variance is larger at the higher of the two temperatures (numerator) when standardized to 10&deg;C. $lnRR_{Q_{10}}$, $lnCVR_{Q_{10}}$ and $lnVR_{Q_{10}}$, as with $Q_{10}$ more generally, all assume that the effect of temperature on physiological rates (or changes in relative variance) is log-linear (see *Supplementary Materials* for further discussion). We test and control for any violations of these assumptions in our analysis (see below).

@@

In addition to acclimation, all our models corrected for the possibility of the log-linearity assumption of our effect sizes being violated ([@fig-concept]; and see Supplementary Materials). We predicted that, if $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ were not strictly log-linear than we would see a decrease in average effect size for studies applying higher temperature treatments because these temperatures are expected to either converge on or cross the thermal optima of the performance curve causing reaction rates to decelerate or decrease beyond $T_{op}$ [@michaletz2024hotter]. The benefit of such an approach means that we could still take advantage of the excellent properties of $Q_{10}$ as an effect size while statistically correcting for the potential non-linearity that would be expected in the data at high treatment temperatures. Given that our data included a wide range of species and habitats, we also included a random slope of maximum temperature that varied across species because we, *a priori*, expected that each species will vary in their thermal performance curve and the chosen temperatures experimentally manipulated along the performance curve. To improve model interpretation, we also mean-centered the maximum temperature and included it in our models.

@@

Changes in relative variance in physiological rates across habitats was nearly equivalent (marine) or larger (terrestrial) compared to their capacity to acclimate. Only in freshwater ectotherms did acclimation in physiological rates coincide with no change in relative variation. Mechanistically, it is unclear what is contributing to the reduction in relative variation in physiological rates at higher temperatures. One possible explanation may be that biochemical reactions are operating more efficiently or are relying on more conformationally stable enzymes at higher temperatures impacting between individual variation [@tattersall2012coping; @angilletta2009thermal; @somero1995proteins]. Such a mechanism may also help explain the differences in relative variance across habitat types. For example, terrestrial ectotherms may rely on more specialised enzymatic variants that operate over a more limited temperature range individually but together allow for organisms to function over a wider thermal range necessary to operate in more variable thermal environments [@angilletta2009thermal]. Acclimation responses in marine and freshwater organisms may also themselves buffer / maintain variance in physiological rates at higher temperatures due to the expression of alternative enzymes that function more effectively at higher temperatures. Given that biochemical traits in marine and freshwater organisms acclimated, and their relative variance increased in freshwater after acclimation, this seems plausible.

@@

### ***Plasticity and relative variance in physiological rates do not differ between life stages***
Acclimation capacities are expected to differ between life-stages because of distinct of patterns of dispersal, habitat use and behaviour that force earlier life stages to cope with more variable environmental conditions which can also lead to developmental constraints on how physiological systems respond later in life [@martin2015age; @stearns1976life; @angilletta2009thermal; @noble2018developmental; @o2019developmental; @Pottier2022]. In addition, plastic responses are also expected to be costly [@angilletta2009thermal; @Dewitt1998], and such costs can be magnified in late life reducing the capacity for plasticity [e.g., @rossi2019ageing]. These processes can also result in changes to intrapopulation variation in physiological rates at higher temperatures but the direction of change between early and adult life stages is likely to depend on the costs of adjusting physiological processes, energy reserves at different life stages, and the extent to which early life experiences constraint plasticity.

Despite these expectations, more broadly, we show little difference between early and late life acclimation capacities and little change in the relative variance in physiological rates across habitats. This may not be too surprising given that such responses are likely context or trait dependent [@carter2020life; @moghadam2019heat]. The lack of differences we observed may also be due to the fact that both juvenile and adult animals occupy similar thermal niches with similar dispersal and thermoregulatory behaviors making physiological responses to temperature similar. It is possible that temperatures experienced earlier in development (e.g., egg or larval stages) exhibit greater capacity for acclimation because of limited dispersal capabilities and very different habitats. Some empirical evidence supports this in insects. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. To date, however, there are few studies that have focused on how variability in responses change and we would encourage more empirical focus on this questions given the potential ecological and evolutionary implications.

### ***No signature of past climate on capacity for physiological plasticity***
Theoretical models predict that plasticity should evolve in populations experiencing greater environmental variability (spatial or temporal), particularly when oscillations are predictable over time to make environmental cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogeneity in terrestrial habitats [@steele2019comparison] might suggest that plasticity should be more likely to evolve in terrestrial environments. However, if thermal variability is too high and unpredictable, the rates of acclimation slow, and there are costs associated with re-modelling physiological processes [which seem likely, @angilletta2009thermal] then we would instead expect canalised responses [@angilletta2009thermal; @Seebacher2015-wi; @loughland2020differences; @Leung2020; @rescan2022experimental; @leung2023phenotypic]. The lack of acclimation in terrestrial ectotherms we observed is consistent with the latter hypothesis, and is supported by other meta-analyses of heat tolerance [@barley2021limited; @Gunderson2015-hz].

However, whether a populations capacity for physiological plasticity and the thermal variability (or predictability) it experiences are related has received more mixed results. We show no relationship between acclimation capacity and thermal variability in marine, freshwater and terrestrial habitats. Our results are consistent with @Gunderson2015-hz who show no relationship between plasticity in heat tolerance and latitude or thermal seasonality. However, other analyses on heat tolerance limits have found relationships between latitude (a proxy for seasonality) [@morley2019physiological] or measures of thermal variability [@verberk2024heat]. @Seebacher2015-wi also found evidence that acclimation capacity is related to a populations thermal variability, however, relationships depended on the habitat and traits in question. Discrepancies across studies could be related to the taxa included in analyses [e.g., @morley2019physiological] or possibly the climate projections/models being used to quantify thermal variability. Modelling realistic microenvironments across such diverse taxa is challenging because it is unclear what the most appropriate spatial and temporal scale might be for specific taxa that is of evolutionary relevance.

@@

Enhanced knowledge of how variation in physiological rates vary across populations and species, and the degree to which they can be adjusted in response to the environment, may lead to more informed predictions about the ecological and evolutionary dynamics of natural populations [@forsman2015rethinking; @seebacher2023can; @cooke2021one; @Sanderson2023]. While we show general patterns across taxa and habitats it is important to recognise that this does not mean such patterns apply to all populations. Substantial variation in acclimation responses and changes in variance exist among populations and traits, as evidenced by wide prediction intervals and substantial species-level and trait-level variance estimates. Nonetheless, quantitative measures on the changes in the relative variance in physiological rates could be better incorporated into physiological and ecological models to provide more nuanced, and possibly more realistic, predictions about the impacts of climate change on natural populations. While we do not yet understand the relative contribution of environmental and genetic factors to variance changes, models could better decouple how different levels of heritability and total variance impact evolutionary and ecological predictions. Our meta-analysis now provides the opportunity to parameterise such models, and ensure they are better aligned with empirical findings.

Many fascinating questions remain unanswered that will require greater focus on the consequences of changes in variance (rather than just the mean). Particularly interesting questions include: How do differences in physiological rate variance change energy flow across trophic levels within communities? What are the biochemical, cellular, and physiological mechanisms that underlie differences in physiological rate variance across habitats? Are reductions in variance in one trait associated with reductions in correlated traits, or do some traits increase while others decrease? How much of the reduction in variance is driven by lower levels of genetic variance? Answers to these questions will require integrative approaches that combine empirical and theoretical work across multiple levels of biological organisation but will likely provide useful advances in understanding the full consequences that climate change will have on ectotherms across major ecosystems globally.

@@

#### *Further discussions on the assumptions of $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates*

$lnRR_{Q_{10}}$, $lnCVR_{Q_{10}}$ and $lnVR_{Q_{10}}$, as with $Q_{10}$ more generally, all assume that the effect of temperature on physiological rates (or changes in relative variance) is log-linear. While this is likely in our data given that we restricted our analysis of $Q_{10}$ to standard operating temperatures for a given species, it may not always be satisfied given the diversity of species in our dataset. $Q_{10}$ [@van1884etudes] has been used exstensively in the physiological literature to successfully address a multitude of questions [e.g., @Havird2020-mj; @Seebacher2015-wi]. However, there is a preference for using a Boltzmann – Arrhenius (BA) relationship (or its more recent extension, the Sharpe-Schoolfield model [@molnar2017thermal; @michaletz2024hotter]) to model thermal effects on physiological rates [@gillooly2001effects; @michaletz2024hotter]. While debate still exists over the utility of $Q_{10}$ when modelling temperature-dependence it is important to recognise that both BA and $Q_{10}$ can exhibit curvilinearity as temperatures increase (as discussed in [@michaletz2024hotter]). @white2012information also showed that the BA model may not always perform better. For example, in eukaryotes, modelling thermal dependence using $Q_{10}$ provided a 5.8-fold better fit to metabolic rate data than the BA relationship [@white2012information]. Given that studies included in our analysis never measured full performance curves at acute and acclimation temperatures it was not possible for us to compare different models of thermal dependence. Nonetheless, $Q_{10}$-based effect sizes remain the most practical effect-size for comparing thermal dependence when using existing empircial data, with the benefit that these effects having convenient properties that make them suitable for meta-analysis. Nonetheless, we control for possible violations of the log-linearity assumption in our analyses.
